---
title: Connect to an SQL Database from RStudio
author: Sunita Kenner
date: '2019-01-20'
slug: connect-to-an-sql-database
categories:
  - ETL
tags:
  - odbc
  - Dbi
description: ''
featured: ''
featuredalt: ''
featuredpath: ''
linktitle: ''
type: post
---

The objective of this post is to supplement and document the process of connecting to an SQL database from RStudio.  

1.  Follow instructions at [rstudio Connect to database](https://db.rstudio.com/getting-started/connect-to-database/).  

```{r, eval = F}

1. brew install freetds:  "--with-unixodbc has depreciated, not needed"
2. brew install psqlodbc
3. brew install mysql
4. brew install sqliteodbc
5. Find odbcinst.ini
   sudo find / -name odbcinst.ini (should be in usr/local/etc/)
   nano /usr/local/etc/odbcinst.ini

6. Add the following & save:

[PostgreSQL Driver]
Driver          = /usr/local/lib/psqlodbcw.so

[SQLite Driver]
Driver          = /usr/local/lib/libsqlite3odbc.dylib

[ODBC for FreeTDS]
Driver          = /usr/local/lib/libtdsodbc.so

7. Find odbc.ini
    sudo find / -name odbc.ini (should be in usr/local/etc/)
    nano /usr/local/etc/odbc.ini

8. Add the following & save

[PostgreSQL]
Driver              = PostgreSQL Driver
Database            = test_db
Servername          = localhost
UserName            = postgres
Password            = password
Port                = 5432

[SQLite]
Driver          = SQLite Driver
Database=/tmp/testing

[ODBC for FreeTDS]
Driver              = ODBC for FreeTDS
Database            = nameofthedatabase
Servername          = nameofserver.com
UserName            = username
Password            = passwd
Port                = Portnumber

9. In R, run the console command to check if drivers are installed. . .
odbcListDrivers()

10. run the console command 
install.packages("odbc")

```

Test it.
```{r, eval = F}
#Load required libraries
library(odbc)
library(DBI)
library(dplyr)

#Define Connection
myconnection <- DBI::dbConnect(odbc::odbc(),
                 driver = "libtdsodbc.so",
                 server = "Server Name",
                 database = "Database name",
                 UID = "userid",
                 Pwd = "passwd",
                 port = 1433)

```

Define an sql chunk and run SQL queries in R-markdown using: **{sql, connection = myconnection, output.var = "nameoftheoutputfile"}**

```{r,  eval = F}

SELECT *, 
      convert(INT, DATENAME(YEAR, DATEADD(month, 6, First_Pur_Date))) FY_FirstPurchase,
      convert(INT, DATENAME(YEAR, DATEADD(month, 6, MaxMED))) FY_EMaintD,
      convert(INT, MONTH(First_Pur_Date)) Month_FirstPurch
FROM "databasename"."tablename"."dataset"

```


Thanks to Mike Griffith, our IT ninja guy, who helped me through this process. 
